DRAFT: Direct-style process creation on Linux

### 2020-05-16 research/paper/paper.tex
paper

Fork-style (such as \texttt{fork}) and spawn-style (such as \texttt{posix\_spawn}) each have advantages and disadvantages.
Correct and performant usage of fork-style places significant constraints on the calling process.
Spawn-style lifts many of these constraints,
but is fundamentally limited in the kinds of processes it can create.
With both fork-style and spawn-style,
it is difficult to
branch based on results or errors encountered while initializing the new process:
fork-style requires IPC if it wishes to communicate errors back to the rest of the program,
and robust indication of errors in spawn-style requires substantial growth in the spawn interface.
Additionally, modern operating systems such as Linux have features
which are difficult or impossible to fully exploit through fork-style or spawn-style.

Direct-style is rare in general, but relatively common on academic operating systems.
\cite{keykos}\cite{sel4}\cite{exokernel}\cite{fuschia}\cite{singularity}
Direct-style involves creating a new child process in a non-running state,
and then calling individual syscalls from the parent to initialize the child.
Direct-style does not constrain the calling process,
easily branches at any step of the initialization,
and can create any kind of process supported by the underlying system.

Our contribution is an implementation of direct-style process creation for Linux,
making use of our work on the rsyscall project.
The rsyscall project develops language-specific ``process-independent'' syscall libraries for Linux,
which replace libc syscall wrappers and allow a program to perform cross-process syscalls.
rsyscall currently has best support for Python,
but can be ported to other languages;
the examples we show in this paper will be in Python,
but generalize easily to other languages.

% TODO this sounds low-content and boring.
% we should probably remove it?
% In our implementation, a parent process calls the \texttt{clone} syscall
% to create a child process which shares memory and other resources with the parent,
% and which runs no user code.
% The parent then uses \texttt{rsyscall}
% to call arbitrary syscalls in the child process to initialize it;
% for example, the parent might call \texttt{chdir} to change the child's current working directory.
% After the process is fully initialized,
% the parent can call \texttt{execve} in the child,
% causing the process to start running an arbitrary executable
% and detaching the child from the control of \texttt{rsyscall}.

Our direct-style process creation implementation uses \texttt{clone} to create a child process in an embryonic state
which can then be manipulated from user code using syscalls with \texttt{rsyscall}.
A direct-style \texttt{clone} has all the advantages of direct-style process creation
over Linux's fork-style \texttt{fork} syscall and spawn-style \texttt{posix\_spawn} libc function.
\begin{itemize}
\item
Direct-style \texttt{clone} can be used correctly and performantly from any process.
\item
Cross-process syscalls made with rsyscall individually report errors, as normal, and can be easily branched on.
\item
Arbitrary processes involving complex namespaces and initialization can be created
using techniques that are impractical with \texttt{fork}.
\item
The performance cost from high-level code varies, but the underlying primitives are substantially faster.
\end{itemize}

We've implemented a variety of useful applications using direct-style process creation on Linux,
which can be expressed more concisely and more clearly than with fork-style,
which we'll showcase in brief in this paper.
Direct-style \texttt{clone} has minimal performance cost in normal situations,
and can outperform \texttt{fork} in some cases.
We've benchmarked our Python implementation of direct-style \texttt{clone}
as comparable to the Python standard library's process creation mechanism.
Finally, we have deployed direct-style process creation in production internally at Two Sigma.

Our implementation is distributed as a feature of \texttt{rsyscall},
and is immediately deployable on recent Linux kernels.
rsyscall is open source and available from \url{https://github.com/catern/rsyscall}.

This paper is organized in several sections.
In section \ref{background}, we provide background on fork-style, spawn-style, and direct-style process creation,
and consider their advantages and disadvantages in more depth.
In section \ref{overview}, we discuss the design of the API,
including our usage of \texttt{rsyscall}.
In section \ref{applications}, we show examples of usage of direct-style process creation in Python on Linux,
including use of process features which are difficult to use on Linux with \texttt{fork} or \texttt{posix\_spawn}.
In section \ref{implementation}, we explore our implementation in depth,
and discuss several difficult process-creation related details of \texttt{rsyscall}'s implementation.
In section \ref{evaluation}, we evaluate, somehow (TODO).
In section \ref{related_work},
we examine other work which also touches on the topics covered by this paper.
In section \ref{future_work},
we list several directions for future work related to direct-style process creation and to \texttt{rsyscall}.
% TODO
In section \ref{conclusions}, we briefly and optimistically conclude.

\section{Background}\label{background}
On most systems today,
the interfaces for process creation
can be divided into "fork-style" and "spawn-style".
A few systems use a third style, which we refer to as "direct-style".
\subsection{Fork-style}
By far the most well-known fork-style interface is Unix's fork.\cite{forkhist}
With a fork-style interface,
a process, called the parent process, calls \texttt{fork},
and a copy of that process is made,
called the child process,
differing only in a few attributes such as process id.
The same program continues executing in both processes,
with a different return value from fork to allow it to distinguish whether it is running in the parent or the child.
Typically, the program will case on the return value
and, if running in the child,
call various system calls to modify the state of the new process.

@@

Multi-threaded programs generally cannot use fork safely.
Typical Unix fork implementations duplicate only the thread calling fork from the parent process to the child process.
In a multi-threaded program, this can cause various issues;
for example, another thread might be holding a memory allocator lock at the time of the fork,
which in the child process will never be unlocked,
causing the child process to deadlock if it tries to allocate memory.
Some thread libraries provide partial mitigations for this issue,
but it's up to user code to make use of those mitigations.\cite{pthread_atfork}
In combination with fork's poor performance in large-memory programs,
a user of fork must think carefully
about the characteristics of the process from which fork is being called.
\subsection{Spawn-style}
In a spawn-style interface,
details about the new process are provided up-front as arguments to some function
which creates the new process all at once.
The most well known spawn-style interfaces are Unix's \texttt{posix\_spawn} \cite{posix_spawn}
and Windows \texttt{CreateProcess} \cite{create_process};
the Unix shell also serves as a spawn-style process creation interface,
especially when used with Bernstein chaining.\cite{chainloading}

@@

Spawn-style interfaces lift the constraints on the calling process that fork-style interfaces impose.
Spawn-style interfaces don't run user code in the new process during initialization.
This means there is no need to copy memory,
so performance is good when a spawn-style interface is used from a process with many memory mappings.
Likewise, there are no inherent issues when a spawn-style interface is used from a multi-threaded program,
though operating systems which implement spawn-style interfaces on top of \texttt{fork} can still have threading bugs.

@@

However, we believe that a natural port of direct-style process creation to Linux
provides an interface that is just as simple to use as fork-style or spawn-style on Linux,
without their disadvantages.
Such an interface, like fork-style and spawn-style,
would implicitly copy some attributes to the new process,
rather than being capability-oriented and requiring all attributes to be specified explicitly.
This is the kind of direct-style interface that we contribute in this paper.

\section{Overview}\label{overview}
In most operating system APIs, including the ones available on Linux,
when a program makes a syscall, it implicitly operates on the current process.
To perform direct-style process creation on Linux,
we need a Linux API where we instead explicitly specify in which process we want to make a syscall.

\texttt{rsyscall} provides this API for Linux.
The \texttt{rsyscall} project develops
language-specific, object-capability-model, low-abstraction libraries for the Linux syscall interface,
bypassing libc.

Unlike POSIX C libraries such as glibc or musl,
an rsyscall library is organized based on the object-capability model.
The capabilities to make syscalls in any given process are reified as language objects.
If the capability to make syscalls in a specific process is not passed (in some way) to a function,
then that function cannot make syscalls in that process.

Due to this design, an \texttt{rsyscall} program can make use of multiple processes at once,
by manipulating capabilities for multiple processes.
The relevant two capabilities for this paper are the initial capability to the ``local'' process,
and capabilities to child processes.
The ``local'' process is the one which hosts the runtime for the running program,
and in which a legacy libc would implicitly make syscalls;
every rsyscall program starts with the ``local'' process
and uses it to bootstrap the initial capabilities for other processes.

We use the term ``thread'' to refer to the processes under the control of rsyscall,
including the ``local'' process.
On Linux, the shared-memory ``threads'' provided by libraries such as \texttt{pthreads}
are implemented as processes,
and their lifetime and execution is completely controlled by a single program.
The same is mostly true of the controlled-process ``threads'' provided by rsyscall.
The most significant difference is that rsyscall ``threads''
do not run user code concurrently with the main program;
nevertheless rsyscall ``threads'' do provide the opportunity for concurrent execution of system calls,
and so the terminology provides useful intuition.

@@

File descriptor passing over a Unix domain socket
can allow similar behavior without sharing the file descriptor table,
but is substantially more complex for the user.
Nevertheless, for fork-style and spawn-style process creation,
file descriptor passing over a Unix domain socket
is the only option to do something like this.
\section{Implementation}\label{implementation}
\subsection{Basics about rsyscall}
In this section, we'll give a brief overview of the implementation of rsyscall,
and then focus on implementation issues specific to process creation.

rsyscall can be conceptually divided in two parts:
the basic cross-process syscall primitive,
and a language-specific library built on top.
The Python language-specific library has already been demonstrated above.
Such libraries only need to be able to call syscalls and explicitly specify a process in some way;
they are, for the most part, agnostic to how the cross-process syscall is implemented.

On Linux \verb|x86_64|, a syscall is specified by a syscall number plus six register-sized arguments;
a syscall returns one register-sized value.
rsyscall's default implementation of cross-process syscalls sends those seven integers over a pipe,
and waits for a response on another pipe.
Processes are created running an infinite loop which, at each iteration,
reads a syscall request off the pipe,
performs that syscall,
and writes the return value back over the return pipe.
In this way, a cross-process syscall works much like a very primitive remote procedure call.

@@

ptrace provides an alternative means to perform arbitrary actions on other processes,
including system calls.
However, among other issues, it has the unavoidable substantial disadvantage of not permitting multiple ptracers.
A ptrace-based implementation would prevent using strace or gdb on rsyscall-controlled processes,
which is an unacceptable limitation for a general-purpose utility.

Many syscalls either take or return pointers to memory,
and require the caller to read or write that memory to provide arguments or receive results.
Therefore, an rsyscall library needs a way to access memory in the target process.

@@

The \verb|process_vm_readv| and \verb|process_vm_writev| system calls
allow the caller to read and write memory from the virtual address space of other processes.
However, they require that the caller have specific credentials relative to the process being accessed,
which may not always be the case.
Additionally, these system calls are disabled if ptrace is disabled system-wide,
which is a niche but possible system configuration.
To ensure that rsyscall can be used for arbitrary purposes and on arbitrary systems, we avoided these calls.
\subsection{\texttt{clone}}\label{clone}
Now that we've established the basic implementation details of rsyscall,
we'll consider specific issues related to process creation and initialization.

@@

\texttt{clone} (along with \texttt{fork}) creates a new process
which immediately starts executing at the next instruction after the syscall instruction,
in parallel with the parent process,
with its registers in generally the same state as the parent process.
\footnote{Note that =glibc= defines a wrapper for the clone syscall;
we are talking about the raw kernel syscall.}
In the style of Plan 9's \texttt{fork} syscall\cite{rfork}, which inspired \texttt{clone},
\texttt{clone} takes a mask of flags which determines whether several attributes of the new process
are either shared with, or copied from, the parent process.

@@

We typically cause the new process to jump to a trampoline provided by the rsyscall library
which sets all registers to values found on the stack
and then jumps to another address.
\footnote{This is also a generally useful utility for hackers performing return-oriented-programming attacks;
but similar functionality exists in any standards-compliant C library,
so there is no increase in attack surface.}
With this trampoline,
we can provide a helper Python function that,
when given a function pointer following C calling conventions, and some arguments,
will prepare a stack for a call to clone such that the new process will call that function with those arguments.

@@

Some systems call system calls in other processes for debugging or introspection purposes.
strace, gdb, and CRIU\cite{criu} are examples of this.
These systems typically uses ptrace,
which is capable of operating on multiple processes at once.
Unfortunately, as discussed in section \ref{implementation},
ptrace is unsuitable as a mechanism for a general purpose system since it is not re-entrant;
a process which is already being ptraced cannot be ptraced again,
so using ptrace as a cross-process system call mechanism would disallow use of strace or debuggers.
\subsection{Process capabilities}
We mentioned the concept of process capabilities while discussing the design of rsyscall in section \ref{overview}.
The Capsicum system provides something called process descriptors\cite{capsicum},
which is a file descriptor handle for a process.
A similar feature has been recently added to Linux in the form of the pidfd API\cite{pidfd}.
These notions of process capability are quite limited, however;
pidfds and process descriptors only allow sending signals to a process or waiting for its death,
rather than exercising full control over the process.
\subsection{Capability-secure libc replacements}
The Capsicum system provides a set of system calls
which can be used to provide a capability-secure sandbox.\cite{capsicum}
Latter efforts\cite{oblivious}, including CloudABI\cite{cloudabi} and WASI\cite{wasi},
are focused primarily on backwards compatibility with the existing POSIX environment, and ease of porting existing code.
Programs ported to the capability-secure environment
are typically launched with an accompanying spawn-style wrapper.
This means that capability-secure processes are created primarily by an API
that is distinct from the capability-secure syscall API they have access to,
which makes it harder for a capability-secure process to spawn further capability-secure processes of its own.
Direct-style process creation could make it easier to create capability-secure processes 
from the capability-secure syscall API,
and reduce the need for a separate spawn-style wrapper.

Also related is the PLASH, ``Principle of Least Authority Shell''\cite{plash}.
PLASH provides a spawn-style API (in the form of a shell)
to launch processes in a capability-secure environment.
PLASH, like all spawn-style APIs, abstracts over the native Linux environment,
and is therefore limited in what kind of processes it can create.
\section{Future work}\label{future_work}
\subsection{Other applications of rsyscall}
Most avenues of future work focus on rsyscall.
rsyscall was not developed solely for the purpose of this paper,
and it has many uses unrelated to direct-style process creation,
such as asynchronous system calls, exceptionless system calls\cite{flexsc}, cross-host operations, among others.
We are actively exploring such applications,
as well as broadening rsyscall's language support.
\subsection{Kernel support}
rsyscall's cross-process syscalls can be performed entirely in userspace,
which has substantial benefits for deployability.
Nevertheless, direct kernel support for creating a stub process and performing syscalls in the context of that process
may provide efficiency benefits, as well as reducing userspace-visible complexity.

Several other aspects of our implementation would be improved by kernel support.
We discussed these in the implementation section;
in brief, we would most benefit from kernel support for
detecting when a child process finishes =execve=,
closing all =CLOEXEC= file descriptors except for an explicitly specified list,
and explicitly specifying a list of =CLOEXEC= file descriptors to inherit over =excve=.
Implementing these features in the kernel in a generally useful way, and upstreaming them,
is an important direction for our future work.
\subsection{Portability to other Unix systems}
Other non-Linux systems
could adopt the techniques of this paper
to provide direct-style process creation.
Currently, our focus is on Linux,
but others may wish to explore porting these techniques to other operating systems.
\subsection{Large scale open source usage}
We have made use of the techniques described in this paper
in proprietary software at Two Sigma.
While this gives us personally greater confidence in these techniques,
it would be better to use them in a publicly available, open source system.
Either porting an existing system to use these techniques,
or using these techniques to create a substantial new system from scratch,
would provide a meaningful demonstration of the viability of these techniques.
% \subsection{file descriptor lifetime management}
% % TODO this is hard to explain, probably best to drop this
% Keeping file descriptors open if and only if there is a specific process using that file descriptor
% is not the only possibility.
% Keep in mind that "a process using the file descriptor" is a slight abuse of terminology;
% processes don't use file descriptors, programs do, and in our system there is only one program.
% To reflect this, we could instead more directly use file descriptors in a process-agnostic way;
% this would support the creation of objects
% which work transparently across multiple processes which share a file descriptor table.
% Such objects would automatically use the relevant process to perform the syscalls for any specific operation.
% A process-centric view of file descriptors instead forces each object to be associated with one process.
% Nevertheless, we found that the process-centric perspective better matches the existing intuitions of users,
% especially those with prior experience in programming with processes.
% We hope that future systems for multi-process programming
% might explore an object-centric approach for managing resources.
\section{Conclusions}\label{conclusions}
Direct-style process creation is much less known and much less used than fork-style and spawn-style.
We have implemented direct-style process creation for Linux.
Our implementation is immediately deployable on today's Linux systems.
We have discussed various applications of processes,
and demonstrated the use of Linux direct-style process creation
to implement them.
We hope that this work will help encourage more use of the process abstraction,
which, though widespread,
is still not used to its full potential.

>>>>

Fork-style (such as \texttt{fork}) and spawn-style (such as \texttt{posix\_spawn}) are the most popular,
but have a number of disadvantages.
Correct, performant use of fork-style places significant constraints on the calling process,
and also has a complex programming model.
Spawn-style lifts many of these constraints, and is simpler to program against,
but is fundamentally limited in the kinds of processes it can create.
With both fork-style and spawn-style,
it is difficult to
branch based on results or errors encountered while initializing the new process:
fork-style requires IPC if it wishes to communicate errors back to the rest of the program,
and robust indication of errors in spawn-style requires substantial growth in the spawn interface.

Direct-style is more rarely seen, but has a number of advantages.
Direct-style involves creating a new child process in a non-running state,
and then calling individual system calls from the parent to initialize the child.
Direct-style does not constrain the calling process,
easily branches at any step of the initialization,
and can create any kind of process supported by the underlying system.

Our contribution is an implementation of direct-style process creation for Linux,
making use of our work on the \texttt{rsyscall} project.
The \texttt{rsyscall} project develops language-specific ``process-independent'' system call libraries for Linux,
which replace libc system call wrappers, and allow a program to perform cross-process system calls.
rsyscall currently has best support for Python,
but can be ported to other languages;
the examples we show in this paper will be in Python,
but generalize easily to other languages.

Our direct-style process creation implementation uses \texttt{clone} to create a child process in an embryonic state
which can then be manipulated from user code using syscalls with \texttt{rsyscall}.
A direct-style \texttt{clone} has all the advantages of direct-style process creation
over Linux's fork-style \texttt{fork} and spawn-style \texttt{posix\_spawn}:
\begin{itemize}
\item
  Processes can be created correctly and performantly from any parent process,
  with a straightforward, familiar API.
\item
  Processes can be created with the full set of Linux features.
\item
  Syscall results are reported individually to the parent process,
  rather than batched together as with spawn-style or manually reported via IPC as with fork-style.
\end{itemize}

We've implemented a variety of useful applications using direct-style process creation on Linux,
which we'll showcase in brief in this paper.
Direct-style \texttt{clone} has minimal performance cost in normal situations,
and can outperform \texttt{fork} in some cases.
We have deployed direct-style process creation in production internally at Two Sigma.

Our implementation is distributed as a feature of \texttt{rsyscall},
and is immediately deployable on recent Linux kernels.
\texttt{rsyscall} is open source and available from \url{https://github.com/catern/rsyscall}.

This paper is organized in several sections.
\begin{itemize}
\item In section \ref{background}, we provide background on fork-style, spawn-style, and direct-style process creation,
and consider their advantages and disadvantages in more depth.
\item In section \ref{overview}, we discuss the design of the API,
including our usage of \texttt{rsyscall}.
\item In section \ref{applications}, we show examples of usage of direct-style process creation in Python on Linux,
including use of process features which are difficult to use on Linux with \texttt{fork} or \texttt{posix\_spawn}.
\item In section \ref{implementation}, we explore our implementation in depth,
and discuss several difficult process-creation related details of \texttt{rsyscall}'s implementation.
\item In section \ref{evaluation}, we evaluate, somehow (TODO).
\item In section \ref{related_work},
we examine other work which also touches on the topics covered by this paper.
\item In section \ref{future_work},
we list several directions for future work related to direct-style process creation and to \texttt{rsyscall}.
\item In section \ref{conclusions}, we briefly outline our conclusions about direct-style and the future of process creation.
\end{itemize}
\section{Background}\label{background}
As mentioned in the introduction,
on most operating systems today,
the available interfaces for process creation
are either "fork-style" or "spawn-style".
A few operating systems have a third style, which we refer to as "direct-style".
\subsection{Fork-style}
By far the most well-known fork-style interface is Unix's fork.\cite{forkhist}
With a fork-style interface,
a process, called the parent process, calls \texttt{fork},
and a copy of that process is made,
called the child process,
differing only in a few attributes such as process id.
The same program continues executing in both processes,
with a different return value from fork to allow it to distinguish whether it is running in the parent or the child.
Typically, the program will case on the return value
and, if running in the child,
call various system calls to modify the state of the new process.

@@

Multi-threaded programs generally cannot use fork safely.
Typical Unix fork implementations duplicate only the thread calling fork from the parent process to the child process.
In a multi-threaded program, this can cause various issues;
for example, another thread might be holding a memory allocator lock at the time of the fork,
which in the child process will never be unlocked,
causing the child process to deadlock if it tries to allocate memory.
Some thread libraries provide partial mitigations for this issue,
but it's up to user code to make use of those mitigations.\cite{pthread_atfork}
In combination with fork's poor performance in large-memory programs,
a user of fork must think carefully
about the characteristics of the process from which fork is being called.
\subsection{Spawn-style}
In a spawn-style interface,
details about the new process are provided up-front as arguments to some function
which creates the new process all at once.
The most well known spawn-style interfaces are Unix's \texttt{posix\_spawn} \cite{posix_spawn}
and Windows \texttt{CreateProcess} \cite{create_process};
the Unix shell also serves as a spawn-style process creation interface,
especially when used with Bernstein chaining\cite{chainloading}.

@@

Spawn-style interfaces lift the constraints on the calling process that fork-style interfaces impose.
Spawn-style interfaces don't run user code in the new process during initialization.
This means there is no need to copy memory,
so performance is good when a spawn-style interface is used from a process with many memory mappings.
Likewise, there are no inherent issues when a spawn-style interface is used from a multi-threaded program,
though operating systems, such as Linux,
which implement spawn-style interfaces on top of \texttt{fork} can still have threading bugs.

@@

However, we believe that a natural port of direct-style process creation to Linux
provides an interface that is just as simple to use as fork-style or spawn-style on Linux,
without their disadvantages.
Such an interface, like fork-style and spawn-style,
would implicitly copy some attributes to the new process,
rather than being capability-oriented and requiring all attributes to be specified explicitly.
This is the kind of direct-style interface that we contribute in this paper.
\section{Overview}\label{overview}
In most operating system APIs, including the ones available on Linux,
when a program makes a syscall, it implicitly operates on the current process.
To perform direct-style process creation on Linux,
we need an API where we instead explicitly specify in which process we want to make a syscall.

\texttt{rsyscall} provides this API.
The \texttt{rsyscall} project develops
language-specific, object-capability-model, low-abstraction libraries for the Linux syscall interface,
bypassing libc.

Unlike POSIX C libraries such as glibc or musl,
an \texttt{rsyscall} library is organized based on the object-capability model.
The capabilities to make syscalls in any given process are reified as language objects.
If the capability to make syscalls in a specific process is not passed (in some way) to a function,
then that function cannot make syscalls in that process.

Due to this design, an \texttt{rsyscall} program can make use of multiple processes at once,
by manipulating capabilities for multiple processes.
The relevant two capabilities for this paper are the initial capability to the ``local'' process,
and capabilities to child processes.
The ``local'' process is the one which hosts the runtime for the running program,
and in which a legacy libc would implicitly make syscalls.
Every \texttt{rsyscall} program starts with the ``local'' process
and uses it to bootstrap the initial capabilities for other processes.

We use the term ``thread'' to refer to the processes under the control of \texttt{rsyscall},
including the ``local'' process.
On Linux, the shared-memory ``threads'' provided by libraries such as \texttt{pthreads}
are implemented as processes,
and their lifetime and execution is completely controlled by a single program.
The same is mostly true of the controlled-process ``threads'' provided by \texttt{rsyscall}.
The most significant difference is that \texttt{rsyscall} ``threads''
do not run user code concurrently with the main program;
nevertheless \texttt{rsyscall} ``threads'' do provide the opportunity for concurrent execution of system calls,
and so the terminology provides useful intuition.

@@

File descriptor passing over a Unix domain socket
can allow similar behavior without sharing the file descriptor table,
but is substantially more complex for the user.
Nevertheless, for fork-style and spawn-style process creation,
file descriptor passing over a Unix domain socket
is the only option to do something like this.
\section{Implementation}\label{implementation}
\subsection{Basics about \texttt{rsyscall}}
In this section, we'll give a brief overview of the implementation of \texttt{rsyscall},
and then focus on implementation issues specific to process creation.

\texttt{rsyscall} can be conceptually divided in two parts:
the basic cross-process syscall primitive,
and a language-specific library built on top.
The Python language-specific library has already been demonstrated above.
Such libraries only need to be able to call syscalls and explicitly specify a process in some way;
they are, for the most part, agnostic to how the cross-process syscall is implemented.

On Linux \verb|x86_64|, a syscall is specified by a syscall number plus six register-sized arguments;
a syscall returns one register-sized value.
\texttt{rsyscall}'s default implementation of cross-process syscalls sends those seven integers over a pipe,
and waits for a response on another pipe.
Processes are created running an infinite loop which, at each iteration,
reads a syscall request off the pipe,
performs that syscall,
and writes the return value back over the return pipe.
In this way, a cross-process syscall works much like a very primitive remote procedure call.

@@

ptrace provides an alternative means to perform arbitrary actions on other processes,
including system calls.
However, among other issues, it has the unavoidable substantial disadvantage of not permitting multiple ptracers.
A ptrace-based implementation would prevent using strace or gdb on \texttt{rsyscall}-controlled processes,
which is an unacceptable limitation for a general-purpose utility.

Many syscalls either take or return pointers to memory,
and require the caller to read or write that memory to provide arguments or receive results.
Therefore, an \texttt{rsyscall} library needs a way to access memory in the target process.

@@

The \verb|process_vm_readv| and \verb|process_vm_writev| system calls
allow the caller to read and write memory from the virtual address space of other processes.
However, they require that the caller have specific credentials relative to the process being accessed,
which may not always be the case.
Additionally, these system calls are disabled if ptrace is disabled system-wide,
which is a niche but possible system configuration.
To ensure that \texttt{rsyscall} can be used for arbitrary purposes and on arbitrary systems, we avoided these calls.
\subsection{\texttt{clone}}\label{clone}
Now that we've established the basic implementation details of \texttt{rsyscall},
we'll consider specific issues related to process creation and initialization.

@@

\texttt{clone} (along with \texttt{fork}) creates a new process
which immediately starts executing at the next instruction after the syscall instruction,
in parallel with the parent process,
with its registers in generally the same state as the parent process.
\footnote{Note that glibc defines a wrapper for the clone syscall;
we are talking about the raw kernel syscall.}
In the style of Plan 9's \texttt{fork} syscall\cite{rfork}, which inspired \texttt{clone},
\texttt{clone} takes a mask of flags which determines whether several attributes of the new process
are either shared with, or copied from, the parent process.

@@

We typically cause the new process to jump to a trampoline provided by the \texttt{rsyscall} library
which sets all registers to values found on the stack
and then jumps to another address.
\footnote{This is also a generally useful utility for hackers performing return-oriented-programming attacks;
but similar functionality exists in any standards-compliant C library,
so there is no increase in attack surface.}
With this trampoline,
we can provide a helper Python function that,
when given a function pointer following C calling conventions, and some arguments,
will prepare a stack for a call to clone such that the new process will call that function with those arguments.

@@

Some systems call system calls in other processes for debugging or introspection purposes.
strace, gdb, and CRIU\cite{criu} are examples of this.
These systems typically uses ptrace,
which is capable of operating on multiple processes at once.
Unfortunately, as discussed in section \ref{implementation},
ptrace is unsuitable as a mechanism for a general purpose system since it is not re-entrant;
a process which is already being ptraced cannot be ptraced again,
so using ptrace as a cross-process system call mechanism would disallow use of strace or debuggers.
\subsection{Process capabilities}
We mentioned the concept of process capabilities while discussing the design of \texttt{rsyscall} in section \ref{overview}.
The Capsicum system provides something called process descriptors\cite{capsicum},
which is a file descriptor handle for a process.
A similar feature has been recently added to Linux in the form of the pidfd API\cite{pidfd}.
These notions of process capability are quite limited, however;
pidfds and process descriptors only allow sending signals to a process or waiting for its death,
rather than exercising full control over the process.
\subsection{Capability-secure libc replacements}
The Capsicum system provides a set of system calls
which can be used to provide a capability-secure sandbox.\cite{capsicum}
Latter efforts\cite{oblivious}, including CloudABI\cite{cloudabi} and WASI\cite{wasi},
are focused primarily on backwards compatibility with the existing POSIX environment, and ease of porting existing code.
Programs ported to the capability-secure environment
are typically launched with an accompanying spawn-style wrapper.
This means that capability-secure processes are created primarily by an API
that is distinct from the capability-secure syscall API they have access to,
which makes it harder for a capability-secure process to spawn further capability-secure processes of its own.
Direct-style process creation could make it easier to create capability-secure processes 
from the capability-secure syscall API,
and reduce the need for a separate spawn-style wrapper.

Also related is the PLASH, ``Principle of Least Authority Shell''\cite{plash}.
PLASH provides a spawn-style API (in the form of a shell)
to launch processes in a capability-secure environment.
PLASH, like all spawn-style APIs, abstracts over the native Linux environment,
and is therefore limited in what kind of processes it can create.
\section{Future work}\label{future_work}
\subsection{Other applications of \texttt{rsyscall}}
Most avenues of future work focus on \texttt{rsyscall}.
\texttt{rsyscall} was not developed solely for the purpose of this paper,
and it has many uses unrelated to direct-style process creation,
such as asynchronous system calls, exceptionless system calls\cite{flexsc}, cross-host operations, among others.
We are actively exploring such applications,
as well as broadening \texttt{rsyscall}'s language support.
\subsection{Kernel support}
\texttt{rsyscall}'s cross-process syscalls can be performed entirely in userspace,
which has substantial benefits for deployability.
Nevertheless, direct kernel support for creating a stub process and performing syscalls in the context of that process
may provide efficiency benefits, as well as reducing userspace-visible complexity.

Several other aspects of our implementation would be improved by kernel support.
We discussed these in the implementation section;
in brief, we would most benefit from kernel support for
detecting when a child process finishes =execve=,
closing all =CLOEXEC= file descriptors except for an explicitly specified list,
and explicitly specifying a list of =CLOEXEC= file descriptors to inherit over =excve=.
Implementing these features in the kernel in a generally useful way, and upstreaming them,
is an important direction for our future work.
\subsection{Portability to other Unix systems}
Other non-Linux systems
could adopt the techniques of this paper
to provide direct-style process creation.
Currently, our focus is on Linux,
but others may wish to explore porting these techniques to other operating systems.
\subsection{Large scale open source usage}
We have made use of the techniques described in this paper
in proprietary software at Two Sigma.
While this gives us personally greater confidence in these techniques,
it would be better to use them in a publicly available, open source system.
Either porting an existing system to use these techniques,
or using these techniques to create a substantial new system from scratch,
would provide a meaningful demonstration of the viability of these techniques.
% \subsection{file descriptor lifetime management}
% % TODO this is hard to explain, probably best to drop this
% Keeping file descriptors open if and only if there is a specific process using that file descriptor
% is not the only possibility.
% Keep in mind that "a process using the file descriptor" is a slight abuse of terminology;
% processes don't use file descriptors, programs do, and in our system there is only one program.
% To reflect this, we could instead more directly use file descriptors in a process-agnostic way;
% this would support the creation of objects
% which work transparently across multiple processes which share a file descriptor table.
% Such objects would automatically use the relevant process to perform the syscalls for any specific operation.
% A process-centric view of file descriptors instead forces each object to be associated with one process.
% Nevertheless, we found that the process-centric perspective better matches the existing intuitions of users,
% especially those with prior experience in programming with processes.
% We hope that future systems for multi-process programming
% might explore an object-centric approach for managing resources.
\section{Conclusions}\label{conclusions}
Direct-style process creation is much less known and much less used than fork-style and spawn-style.
We have implemented direct-style process creation for Linux.
Our implementation is immediately deployable on today's Linux systems.
We have discussed various applications of processes,
and demonstrated the use of Linux direct-style process creation
to implement them.

We believe that direct-style process creation is superior to fork-style and spawn-style,
including on Linux.
Our technique is immediately deployable and can be used for a wide variety of complex applications.

We hope that a better process creation mechanism
will help encourage more use of the process abstraction,
which, though widespread,
is still not used to its full potential.
